Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-07 Manuscript/manuscript.tex

The ion concentrations with constant value at the simulation box edges
are also used to quantify the ion binding affinities is different simulation
models by using the relative surface excess of ions with respect to water $\Gamma_i^w$.
This is defined as \cite{chattorajBOOK}
\begin{eqnarray}
  \Gamma_i^w   &=& \Gamma_i-\Gamma_w \frac{C_i-C_i'}{C_w-C_w'} \\
  \Gamma_{a} &=& \frac{n_{a} - C_{a} V - C_{a}' V'}{A} \, \, ; \, a=i,w \, ,
\end{eqnarray}
where index $i$ and $w$ denote ions and water respectively,
prime accent denotes the phase inside bilayer, 
whereas non-prime symbols account for bulk water phase,
$n$ is the total amount of water or ions,
$C$ denotes bulk concentrations, 
$V$ is the volume of the respective phase separated by the Gibbs divinding plane, 
$A$ is the area of the interface, 
and $\Gamma_a$ is the surface excess of water or ions. 
As $\Gamma_i^w$ does not depend on the definition of the interfacial region
and the position of the Gibbs dividing plane~\cite{chattorajBOOK},
we consider the whole simulation box as an interface. 
Note that the area of the interface is given by the simulation box area
multiplied with two $2A_b$, because bilayers have a interface
in both leaflefts.
The concentrations of water and ions inside
the bilayer can be expected to vanish, thus $c_{i,w}'=0$.
This provides us a simplified relation for 
the relative surface excess of lipid bilayers in simulations
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where the bulk concentrations, $C_w$ and $C_i$, are 
concentrations with constant value at the simulation box edges.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.0cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{There seems to a typo in experimental value for other order parameter for C2 carbon in sn-2 chain; it should be negative.
    The scale of the y-axis should be then zoomed when this is fixed.} \\
  \todo{I would put the acyl chain order parameters also for original Lipid14}\\
  \todo{x-axis scale in form factor figure would be better from 0 to 0.6. The experimental data is available for this range.} \\
  \todo{$S_{CD} -> S_{CH}$}
\end{figure}

@@

The headgroup order parameter changes with increasing amount of 
cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the head group order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

@@

The binding affinity of aqueous cations in lipid 
bilayers can be measured by using the headgroup 
order parameters, because they decrease linearly 
with the bound positive charge~\cite{seelig87,catte16}. 
The headgroup order parameter responses 
to aqueous NaCl and CaCl$_2$ concentrations 
are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}
from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC

@@

Negligible changes of the headgroup order parameters are measured
with submolar concentrations of NaCl due to the very low affinity of Na$^+$
in PC bilayers \cite{akutsu81}. While Na$^+$ binding and the related 
headgroup order parameter changes were overestimated in
almost all the available simulation models, the low affinity
and negligible order parameter changes were reproduced by Lipid14 
model when simulated with \AA{}qvist ions~\cite{catte16}.
However, the same combination of force field parameters
overestimated the headgroup order parameter response to CaCl$_2$ 
concentration, which was the case also in all other models tested in
Ref.~\citenum{catte16}.
Using ion model by Dang et al.~\cite{smith94,chang1999,dang2006} or
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
with more realistic bulk behaviour did not improve the results
for the CaCl$_2$ interactions with Lipid14 model, as seen in 
in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively.
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
The results support the conclusion of the previous work \cite{catte16}
that improvements also in lipid models are needed to 
correctly describe cation binding in PC bilayers.

@@

The binding affinities are quantified by using
the ion density profiles along membrane normal shown
in Fig.~\ref{fig:cacl-dens}. The density profiles show
larger Ca$^{2+}$ density peak in lipid headgroup region for
Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
The relative surface excess calculated from Eq.~\ref{surfexcess} gives 
$\Gamma_i^w = 0.07 nm^{-3}$ for the ECC-lipid model,
which is significantly smaller than $\Gamma_i^w = XX nm^{-3}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 nm^{-3}$ with Dang ions.
\todo{Surface excess for Lipid14 with Aqvist yet to be added.} \\
\todo{Surface excess units are not correct. Maybe the numbers needs to be checked also?} \\
\todo{JOE: Add error estimates to Gamma values}
Rough estimates for the free energy difference between bound and unbound cation
are given by
\begin{equation}
  \Delta G={\rm k_b T}\log(\frac{p_o}{p_i}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities2.eps}
  \caption{\label{fig:cacl-dens}
    Number density of \ce{Ca^{2+}} and \ce{Cl^-} as a function of membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    Densities of~\ce{Cl^-} and water are divided with 2 and 200, respectively, to visalize them
    with the same scale as \ce{Ca^{2+}}. The molar concentration of the ions in water is 350~mM in all systems
    presented here. 
    }
  \todo{Calculate the surface excess for L14+Aqvist sim.} \\
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
  \todo{The current data for Dang simulation seems to contain more ions than others.}
\end{figure}

@@

>>>>

The ion concentrations with constant value at the simulation box edges
are also used to quantify the ion binding affinities is different simulation
models by using the relative surface excess of ions with respect to water $\Gamma_i^w$.
This is defined as \cite{chattorajBOOK}
\begin{eqnarray}
  \Gamma_i^w   &=& \Gamma_i-\Gamma_w \frac{C_i-C_i'}{C_w-C_w'} \\
  \Gamma_{a} &=& \frac{n_{a} - C_{a} V - C_{a}' V'}{A} \, \, ; \, a=i,w \, ,
\end{eqnarray}
where index $i$ and $w$ denote ions and water respectively,
prime accent denotes the phase inside bilayer, 
whereas non-prime symbols account for bulk water phase,
$n$ is the total amount of water or ions,
$C$ denotes bulk concentrations, 
$V$ is the volume of the respective phase separated by the Gibbs divinding plane, 
$A$ is the area of the interface, 
and $\Gamma_a$ is the surface excess of water or ions. 
As $\Gamma_i^w$ does not depend on the definition of the interfacial region
and the position of the Gibbs dividing plane~\cite{chattorajBOOK},
we consider the whole simulation box as an interface. 
Because bilayers have an interface
in both leaflefts,
the total area of the interface in the simulation
 is given by the simulation box area multiplied by two, $2A_b$.
The concentrations of water and ions inside
the bilayer can be expected to vanish, thus $c_{i,w}'=0$.
This provides us a simplified relation for 
the relative surface excess of lipid bilayers in simulations
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where the bulk concentrations, $C_w$ and $C_i$, are 
concentrations with constant value at the simulation box edges.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.0cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{There seems to a typo in experimental value for other order parameter for C2 carbon in sn-2 chain; it should be negative.
    The scale of the y-axis should be then zoomed when this is fixed.} \\
  \todo{I would put the acyl chain order parameters also for original Lipid14; JOE: Already done, check whether your figures are up to date. I usually push only *.eps.}\\
  \todo{x-axis scale in form factor figure would be better from 0 to 0.6. The experimental data is available for this range.} \\
  \todo{$S_{CD} -> S_{CH}$}
\end{figure}

@@

The headgroup order parameter changes with increasing amount of 
cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease of the order parameters is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the head group order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

@@

The binding affinity of aqueous cations in lipid 
bilayers can be measured by using the headgroup 
order parameters, because they decrease linearly 
with the bound positive charge~\cite{seelig87,catte16}. 
The headgroup order parameter responses 
to aqueous NaCl and CaCl$_2$ concentrations 
are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}
from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC.

@@

Negligible changes of the headgroup order parameters are measured
with submolar concentrations of NaCl due to the very low affinity of Na$^+$
in PC bilayers \cite{akutsu81}. While Na$^+$ binding and the related 
headgroup order parameter changes were overestimated in
almost all the available simulation models, the low affinity
and negligible order parameter changes were reproduced by Lipid14 
model when simulated with \AA{}qvist ions~\cite{catte16}.
However, the same combination of force field parameters
overestimated the headgroup order parameter response to CaCl$_2$ 
concentration, which was the case also in all other models tested in
Ref.~\citenum{catte16}.
Using ion model by Dang et al.~\cite{smith94,chang1999,dang2006} or
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
with more realistic bulk behaviour did not improve the results
for the CaCl$_2$ interactions with Lipid14 model, as seen 
in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively.
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
The results support the conclusion of the previous work \cite{catte16}
that improvements also in lipid models are needed to 
correctly describe cation binding in PC bilayers.

@@

The binding affinities are quantified by using
the ion density profiles along membrane normal shown
in Fig.~\ref{fig:cacl-dens}. The density profiles show
larger Ca$^{2+}$ density peak in lipid headgroup region for
Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
The relative surface excess calculated from Eq.~\ref{surfexcess} gives 
$\Gamma_i^w = 0.07 \pm 0.01 nm^{-2}$ for the ECC-lipid model,
which is significantly smaller than $\Gamma_i^w = 0.13 \pm 0.01 nm^{-2}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 \pm 0.03 nm^{-2}$ with Dang ions.
Rough estimates for the free energy difference between bound and unbound cation
are given by
\begin{equation}
  \Delta G={\rm k_b T}\log(\frac{p_o}{p_i}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities2.eps}
  \caption{\label{fig:cacl-dens}
    Number density of \ce{Ca^{2+}} and \ce{Cl^-} as a function of membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    Densities of~\ce{Cl^-} and water are divided with 2 and 200, respectively, to visalize them
    with the same scale as \ce{Ca^{2+}}. The molar concentration of the ions in water is 350~mM in all systems
    presented here. 
    }
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
  \todo{The current data for Dang simulation seems to contain more ions than others.}
\end{figure}

@@

%\begin{figure}[tbp]
%  \centering
%  \includegraphics[height=6.0cm]{../Fig/POPC_binding_labels.eps} 
%  \includegraphics[height=6.0cm]{../Fig/POPC_binding_ECC-lipids.eps}
%  \caption{\label{fig:binding_conf}
%      Configurations of \ce{Ca2+} binding in POPC membranes in the ECC-lipids model.
%      The binding sites are: 
%      red circle for phosphate oxygens; 
%      blue square for sn-1 carbony oxygen; and green triangle for sn-2 carbonyl oxygen.
%      The configurations along with their probabilities are drawn in the following manner:
%      each row corresponds to an individual lipid the cation is bound to;
%      symbols denote the site the cations is bound to the respective lipid.
%      An empty configuration denotes an unbound cation in a solvent. 
%  }
%\end{figure}
